---
title: "p8105_hw2_hct2116"
author: "Haoyun Tong"
date: "2026-10-06"
output: github_document
---

```{r setup}
library(tidyverse)
library(readxl)
library(dplyr)
```

## Problem 1

### import, view, clean, convert
```{r}
nyc_transit = read.csv(file = "./hw2 data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv" ) %>%
                janitor::clean_names() %>% 
                select(line, station_name, station_latitude, station_longitude, starts_with("route"),
                       entry, vending, entrance_type, ada) %>% 
                mutate(entry = case_match(entry, "YES" ~ TRUE, "NO" ~ FALSE))
view (nyc_transit)
```

### a short paragraph about dataset and answers to questions
The dataset records NYC subway entrance and exit locations, including the subway line and station name, coordinates, routes served, whether entry is allowed, whether vending is available, entrance type, and ADA compliance. I cleaned the column names first, selected the variables that are required by the assignment, and converted entry from "YES"/"NO" text to a logical variable. The resulting dataset has 19 total columns and 1868 rows. The data are tidy.

```{r}
nyc_transit %>%
  distinct(line, station_name) %>%
  nrow()
```
There are 465 distinct stations.

```{r}
nyc_transit %>%
  filter(ada == TRUE) %>%
  distinct(line, station_name) %>%
  nrow()
```
There are 84 stations that are ADA compliant.

```{r}
nyc_transit %>%
  filter(vending == "NO") %>%
  summarize(proportion = mean(entry, na.rm = TRUE))
```
The proportion is 0.38.

### reformat
```{r}
#reformat
nyc_routes <- nyc_transit %>%
  mutate(across(starts_with("route"), as.character)) %>%
  pivot_longer(
    cols = starts_with("route"),
    names_to = "route_number",
    values_to = "route_name"
  ) %>%
  filter(!is.na(route_name), route_name != "")

# distinct stations serving A train
a_stations <- nyc_routes %>%
  filter(route_name == "A") %>%
  distinct(line, station_name)
nrow(a_stations)

# ada compliant A train stations
nyc_routes %>%
  filter(route_name == "A", ada == TRUE) %>%
  distinct(line, station_name) %>%
nrow()
```
There are 60 distinct stations serve the A train and 17 of them are ADA compliant.


## Problem 2
### import, read, clean (Mr. Trash Wheel)
```{r}
mr_tw <- read_excel("./hw2 data/202610 Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel", range = "A2:N743") %>% 
  janitor::clean_names() %>% 
  filter(!is.na(dumpster)) %>% 
  mutate(sports_balls = as.integer(round(sports_balls)))
view(mr_tw)
```

### import, read, clean (Professor Trash Wheel)
```{r}
prof_tw <- read_excel("./hw2 data/202610 Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel", range = "A2:M140") %>% 
  janitor::clean_names() %>% 
  filter(!is.na(dumpster))
view(prof_tw)
```

### import, read, clean (Gwynnda)
```{r}
gwyd <- read_excel("./hw2 data/202610 Trash Wheel Collection Data.xlsx", sheet = "Gwynnda the Good Wheel of the W", range = "A2:L395") %>% 
  janitor::clean_names() %>% 
  filter(!is.na(dumpster))
view(gwyd)
```

### combine
```{r}
mr_tw <- mr_tw %>% mutate(year = as.integer(year), wheel = "Mr. Trash Wheel")
prof_tw <- prof_tw %>%  mutate(plastic_bottles = as.integer(plastic_bottles), wheel = "Professor Trash Wheel")
gwyd <- gwyd %>%  mutate(wheel = "Gwynnda")

three_trash_wheels <- bind_rows(mr_tw, prof_tw, gwyd) %>% 
  relocate(wheel)
view (three_trash_wheels)
```

### observation with inline R
The combined dataset contains `r nrow(three_trash_wheels)` dumpster observations from Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda. Each observation includes variables such as the wheel, dumpster number, collection date, weight, and numbers of the items collected (for example, cigarette butts and sports balls). Professor Trash Wheel collected `r three_trash_wheels %>% filter(wheel == "Professor Trash Wheel") %>% summarize(total = sum(weight_tons, na.rm = TRUE)) %>% pull(total)` tons of trash in total. Gwynnda collected `r three_trash_wheels %>% filter(wheel == "Gwynnda", month == "June", year == 2022) %>% summarize(total = sum(cigarette_butts, na.rm = TRUE)) %>% pull(total) %>% format(big.mark = ",", scientific = FALSE)` cigarette butts in June 2022.

## Problem 3
### import, tidy, clean baseline dataset
```{r}
baseline <- read_csv("./hw2 data/data_mci/MCI_baseline.csv",
                     skip = 1,
                     na = ".") %>% 
            janitor::clean_names() %>% 
  mutate(
    sex = case_match(sex, 1 ~ "Male", 0 ~ "Female"),
    apoe4 = case_match(apoe4, 1 ~ TRUE, 0 ~ FALSE)
  )
```
### discussion, summary by inline R
This dataset has a descriptive row before the actual column names, so `skip = 1` skips that row. I also used `na = "."` to treat missing MCI onset ages as missing values. After cleaning the column names, `sex` and `apoe4` are recorded as categories. Using `case_match` converts them into non-numeric variables. There were 483 participants recruited in the study, among them `r sum(!is.na(baseline$age_at_onset))` developed MCI during follow-up. The mean baseline age was `r mean(baseline$current_age) %>% signif(5)` years. Of the `r sum(baseline$sex == "Female")` women, `r sum(baseline$sex == "Female"&baseline$apoe4 == TRUE)` were APOE4 carriers, so the proportion was `r mean(baseline$apoe4[baseline$sex == "Female"])%>% signif(4)`.

### same for the dataset of longitudinally observed biomarker values
```{r}
amyloid <- read_csv("./hw2 data/data_mci/mci_amyloid.csv", skip = 1) %>%
  janitor::clean_names() %>% 
  pivot_longer(cols = baseline:time_8,
               names_to = "visit_time",
               values_to = "amyloid_ratio") %>% 
  mutate(visit_time_year = case_match(
    visit_time,
    "baseline" ~ 0,
    "time_2" ~ 2,
    "time_4"~ 4,
    "time_6"~ 6,
    "time_8"~ 8
  )) %>% 
  relocate(visit_time_year, .before = amyloid_ratio) %>% 
  select(study_id, visit_time_year, amyloid_ratio)
```
When importing this dataset, I also skipped the first row containing descriptions by `skip = 1`. After cleaning the names, I used pivot to transform the wide format into a long format, separating each `visit_time` into different rows. And then I created a new column called `visit_time_year` to store the time in numeric format (eg, baseline as 0, time_2 as 2). To have a clearer final view, I selected the three columns to display: `study_id`, `visit_time_year`, `amyloid_ratio`.

### participant check, combine, describe, export
```{r}
only_baseline <- anti_join(baseline, amyloid, by = c("id" = "study_id"))
nrow(only_baseline)
```
There are 8 baseline-only participants.
```{r}
only_amyloid <- anti_join(amyloid, baseline, by = c("study_id" = "id"))
nrow(only_amyloid)
```
There are 60 amyloid-only participants. 
```{r}
mci_combined <- inner_join(amyloid, baseline, by = c("study_id" = "id"))
view (mci_combined)
```
The inner join will retain the 475 participants present in both. The combined dataset contains only participants present in both the baseline and amyloid datasets. It has one row per participant per amyloid visit. Each row includes the study ID, visit time, amyloid 42/40 ratio, and the participant’s baseline demographic information.
```{r}
write_csv(mci_combined, "mci_combined.csv")
```

